\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-11-07 k-planar.tex
A bunch of changes

The remainder of this paper is organized as follows:
\begin{compactitem}
  \item In \cref{Structure} we prove our main result for shortcut systems (\cref{ShortcutProduct}), and its optimisations for $k$-planar graphs and their genus-$g$ generalisation (\cref{kPlanarProduct,gkPlanarProduct}).
  \item \Cref{examples} uses \cref{ShortcutProduct,kPlanarProduct,gkPlanarProduct} to derive product structure theorems for string graphs, powers of bounded-degree graphs, map graphs, and $k$-nearest neighbour graphs.
  \item \Cref{FramedSection} proves a product structure theorem for $d$-framed multigraphs in which the graph $H$ has treewidth $3$ and is planar.  Using this result, we derive improved product structure theorems for $1$-planar graphs and planar map graphs, again with a planar graph $H$.
  \item \Cref{Applications} quickly surveys applications of product structure theorems and the consequences of the current work.
\end{compactitem}

@@

We now prove \cref{ShortcutPartition}.
For convenience, it will be helpful to assume that $\SS$ contains a length-1 $vw$-shortcut for every edge $vw\in E(G)$.  Since $G^\SS$ is defined to be a supergraph of $G$, this assumption has no effect on $G^{\SS}$ but eliminates special cases in some of our proofs.

@@

\begin{proof}
  Recall that $S_x$ is defined by vertices that participate in $x$, and these are vertices that are either in $Y_x$ or in shortcuts that cross $Y_x$.  We say that a vertex $w\in Y_x$ \defin{contributes} a vertex $v\in S_x$ if $v=w$ or if some path in $\SS$ that contains $v$ has $w$ as an internal vertex.
  We upper bound the number of vertices in $S_x\cap L_i$ by upper-bounding the number of vertices contributed to $S_x\cap L_i$ by each $w\in Y_x$.

@@

\begin{proof}
  For the tree-decomposition $(C_x:x\in V(T))$ of $J$ we use the same tree $T$ used in the tree-decomposition $(B_x:x\in V(T))$ of $H$. For each node $x$ of $T$, we define $C_x$ as follows: $C_x$ contains $x$ as well as every $T$-ancestor $a$ of $x$ such that $J$ contains an edge $ax'$ where $x$ is a $T$-ancestor of $x'$ (including the possibility that $x=x'$).
  \cref{i-ancestor} ensures that, for every edge $ax'\in E(J)$, both $a$ and $x'$ are contained in $C_{x'}$.  The connectivity of $T[a]:=T[\{x\in V(T):a\in C_x\}]$ follows from the fact that, for every node $x'\in V(T[a])$, every node $x$ on the path in $T$ from $a$ to  $x'$ is also a node of $T[a]$.  Therefore $(C_x:x\in V(T))$ is indeed a tree-decomposition of $J$.

@@

\cref{d_framed_product_stucture} is a consequence of the following technical lemma, which is an extension of the analogous result for plane graphs~\cite{DJMMUW20}. The following definition is a convenient way to establish the planarity of $H$ in \cref{d_framed_product_stucture}:  A $T$-decomposition of a graph $G$ is \defin{$t$-simple} if each of its bags has size at most $t+1$ and, for each $t$-element subset $\{v_1,\ldots,v_t\}\subseteq V(G)$, at most two bags contain $v_1,\ldots,v_t$.
To show the planarity of $H$ in \cref{d_framed_product_stucture} we use the fact that a graph has a 3-simple tree-decomposition if and only if it has treewidth at most 3 and is planar~\citep{kratochvil.vaner:note}. Finally, a \defin{BFS spanning forest} $T$ of $G$ rooted at a set $V_0\subseteq V(G)$ is a spanning forest of $G$ with the property that for each $v\in V(G)$,  \[\min\{\dist_G(v,w):w\in V_0\}=\min\{\dist_T(v,w):w\in V_0\}.\]

@@

\begin{proof}
This proof is similar to the proof of Lemma~13 by \citet{DJMMUW20}, but is complicated by several factors.  In particular, the reader should keep in mind that $G$ and $G_0$ are multigraphs and that the cycle $F$ may consist of two vertices and two (parallel) edges.

@@

We now move onto the case in which $N_0$ contains at least one inner vertex.
Our $H$-partition $\mathcal{P}$ will include $P_1,\dots,P_k$ and a (possibly empty) part $S$ that will be determined by three vertices $v_1v_2v_3$ that belong to a common inner face $\tau$ of $N_0$.  We first explain how $v_1$, $v_2$ and $v_3$ are chosen and then explain how these are used to define $S$.  There is one case to consider for each possible value of $k$ (see \cref{boring_figure}):

@@

		\item If $k=3$ then assign each vertex $v$ of $N_0$ a colour $\alpha(v)\in\{1,\ldots,3\}$ as follows:  Let $w$ be the first vertex of $V(F)$ encountered on the path in $T$ from $v$ to the appropriate vertex of $F_0$.  (This implies that $w=v$ when $v$ is a vertex of $F$.)  Since $w\in V(F)$, $w\in P_i$ for some $i\in\{1,\ldots,k\}$ and we define $\alpha(v):=i$.  Now, for each $d'\in\{4,\ldots,d\}$ and each $d'$-sided inner face $F$ of $N_0$, add $d'-3$ edges inside of $F$ to split it into $3$-sided faces.  This yields a plane supergraph $N_0'$ of $N_0$ in which each inner face has exactly three edges on its boundary. By Sperner's Lemma (see \citep{Proofs4}) there exists an inner face $\tau':=v_1,v_2,v_3$ of $N_0'$ such that $\alpha(v_i)=i$ for each $i\in\{1,2,3\}$.  Let $\tau:=v_1,\ldots,v_p$ be  the facial cycle of $N_0$ that contains $\tau'$.
	\end{compactenum}

@@

	We now define the vertices in $S$.  For each $i\in\{1,2,3\}$, let $Q_i$ be the shortest path, in $T$, from $v_i$ to a vertex in $V(F)$.  Let $\overline{S}$ denote the subgraph of $N_0$ consisting of vertices and edges of $Q_1$, $Q_2$, $Q_3$, and $\tau$.
	Let $S:=V(\overline{S})\setminus V(F)$, let $Y:=(V(Q_1)\cup V(Q_2)\cup V(Q_3))\setminus V(F)$, and let $X:=S\setminus Y$. Observe that $|V(Q_i)\cap L_j|\le 1$ 	for each $i\in\{1,2,3\}$ and each $j\ge 0$.  Therefore $|Y\cap L_j|\le 3$ for each $j\ge 0$. Furthermore, $X\subseteq V(\tau)\setminus\{x_1,x_2,x_3\}$, so $|X|\le p-3\le d-3$. Therefore, using $S:=Y\cup X$ as a part in the partition $\PP$ satisfies condition~(ii).

@@

\begin{proof}[Proof of \cref{d_framed_product_stucture}]
Let $G$ be a $d$-framed multigraph with frame $G_0$ having outer face $F_0$
Let $T$ be a BFS forest of $G_0$ rooted at $V(F_0)$ and let $\langle L_0,L_1,\ldots\rangle$ be the resulting BFS layering.  Let $P_1:= V(F_0)$, let $Y_1$ be any subset of $\min\{|P_1|,3\}$ vertices in $P_1$ and let $X_1:=P_1\setminus Y_1$.  Now apply \cref{induction} on $G$, $G_0$, $T$, $F=F_0$, $k=1$, and $P_1=Y_1\cup X_1$.
This gives an $H$-partition $\mathcal{P}:=\{S_x:x\in V(H)\}$ of $G$ in which $H$ has a $3$-simple tree-decomposition.  Since any graph having a $3$-simple tree-decomposition is planar \cite{knauer.ueckerdt:simple,kratochvil.vaner:note,elmallah.colbourn:on}, $H$ is planar.

Define $\mathcal{L}=\langle L_0',L_1'\ldots\rangle$ where $L_i'=L_{\lfloor d/2\rfloor i}\cup \cdots \cup L_{\lfloor d/2\rfloor(i+1)-1}$ for each integer $i\ge 0$. This is a layering of $G$ since $\dist_{G_0}(v,w)\le \floor{ \frac{d}{2} }$ for each edge $vw\in E(G)$.  Since $|L_j\cap Y_x|\le 3$ for each integer $j\ge 0$, $|L_i'\cap Y_x|\le 3\floor{\frac{d}{2}}$ and therefore $|L_i'\cap S_x|= |L_i'\cap Y_x|+ |L_i'\cap X_x|\le \floor{ \frac{d}{2}} + d -3$ for each integer $i\ge 0$ and each $x\in V(H)$. The result now follows from \cref{PartitionProduct}.
\end{proof}

>>>>

The remainder of this paper is organized as follows:
\begin{compactitem}
  \item In \cref{Structure} we prove our main result for shortcut systems (\cref{ShortcutProduct}), and its optimisations for $k$-planar graphs and their genus-$g$ generalisation (\cref{kPlanarProduct,gkPlanarProduct}).
  \item \Cref{examples} uses \cref{ShortcutProduct,kPlanarProduct,gkPlanarProduct} to derive product structure theorems for bounded-degree string graphs, powers of bounded-degree graphs, map graphs, and $k$-nearest neighbour graphs.
  \item \Cref{FramedSection} proves a product structure theorem for $d$-framed multigraphs in which the graph $H$ has treewidth $3$ and is planar.  Using this result, we derive improved product structure theorems for $1$-planar graphs and planar map graphs, again with a planar graph $H$.
  \item \Cref{Applications} quickly surveys applications of product structure theorems and the consequences of the current work.
\end{compactitem}

@@

We now prove \cref{ShortcutPartition}.
For convenience, it will be helpful to assume that $\SS$ contains a length-1 $vw$-shortcut for every edge $vw\in E(G)$.  Since $G^\SS$ is defined to be a supergraph of $G$, this assumption has no effect on $G^{\SS}$ but eliminates special cases in some of our proofs.  For a given $H$-partition $\mathcal{H}:=\{B_x:x\in V(H)\}$ of a graph $G$, we are frequently interested in the unique vertex $x\in V(H)$ such that $B_x$ contains a particular vertex $v\in V(G)$.  We will write this concisely by writing that ``$P_x$ is the part in $\mathcal{H}$ that contains $v$.''

@@

\begin{proof}
  % Recall that $S_x$ is defined by vertices that participate in $x$, and these are vertices that are either in $Y_x$ or in shortcuts that cross $Y_x$.
  It follows from \cref{short_path_in_h} that, for each $v\in V(H)$, $v$ participates in $x:=a(v)$.  Therefore, the contents of $S_x$ are limited to elements that are in $Y_x$ or vertices in shortcuts that cross $x$.  We say that a vertex $w\in Y_x$ \defin{contributes} a vertex $v\in S_x$ if $v=w$ or if some path in $\SS$ that contains $v$ has $w$ as an internal vertex.  We upper bound the number of vertices in $S_x\cap L_i$ by upper-bounding the number of vertices contributed to $S_x\cap L_i$ by each $w\in Y_x$.

@@

\begin{proof}
  For the tree-decomposition $(C_x:x\in V(T))$ of $J$ we use the same tree $T$ used in the tree-decomposition $(B_x:x\in V(T))$ of $H$. For each node $x$ of $T$, we define $C_x$ as follows: $C_x$ contains every $T$-ancestor $a$ of $x$ such that $J$ contains an edge $ax'$ where $x$ is a $T$-ancestor of $x'$ (including the possibility that $x=x'$).  Additionally, $C_x$ includes $x$ if $x\in V(J)$.
  \cref{i-ancestor} ensures that, for every edge $ax'\in E(J)$, both $a$ and $x'$ are contained in $C_{x'}$.  The connectivity of $T[a]:=T[\{x\in V(T):a\in C_x\}]$ follows from the fact that, for every node $x'\in V(T[a])$, every node $x$ on the path in $T$ from $a$ to  $x'$ is also a node of $T[a]$. Therefore $(C_x:x\in V(T))$ is indeed a tree-decomposition of $J$.

@@

\cref{d_framed_product_stucture} is a consequence of the following technical lemma, which is an extension of the analogous result for plane graphs~\cite[Lemma~17]{DJMMUW20}. The following definition is a convenient way to establish the planarity of $H$ in \cref{d_framed_product_stucture}:  A $T$-decomposition of a graph $G$ is \defin{$t$-simple} if each of its bags has size at most $t+1$ and, for each $t$-element subset $\{v_1,\ldots,v_t\}\subseteq V(G)$, at most two bags contain $v_1,\ldots,v_t$.  To show the planarity of $H$ in \cref{d_framed_product_stucture} we use the fact that a graph has a 3-simple tree-decomposition if and only if it has treewidth at most 3 and is planar~\citep{kratochvil.vaner:note}. Finally, a \defin{BFS spanning forest} $T$ of $G$ rooted at a set $V_0\subseteq V(G)$ is a spanning forest of $G$ that contains a tree rooted at each vertex of $V_0$ with the property that for each $v\in V(G)$,
\[  \min\{\dist_G(v,w):w\in V_0\}=\min\{\dist_T(v,w):w\in V_0\}.  \]

% The following technical lemma is analogous to Lemma~17 in \cite{dujmovic.joret.ea:planar}, which performs the inductive step in the construction of a so-called \emph{tripod decomposition}.
%
%
%
%
%
%  cycle $F$ in a near-triangulation $G$ and a partition of $V(F)$ into at most three bipods,
%
%
% to how a near-triangulation bounded by a cycle $F$ in in a triangulation $G$ whose vertices have already been assigned
%
% we offer some comparison.  \cite[Lemma~17]{dujmovic.joret.ea:planar} shows that, given a partial partition of $V(G)$ into tripods, and a cycle $F$ in $G$ whose vert

For readers already familiar with the proof in \cite{dujmovic.joret.ea:planar}, the main difference here is that a typical tripod in the proof of \cite[Lemma~17]{dujmovic.joret.ea:planar} consists of three upward paths in a BFS tree/forest whose lower endpoints belong to a single triangular face $\tau$.  In the current setting, a typical tripod consists of three upward paths belonging to a common face $\tau$ of $G_0$ plus all of the (at most $d$) vertices of $\tau$.

@@

\begin{proof}
This proof is similar to the proofs of Lemmas~13 and 17 in \cite{DJMMUW20}, but is complicated by several factors.  In particular, the reader should keep in mind that $G$ and $G_0$ are multigraphs and that the cycle $F$ may consist of two vertices and two (parallel) edges.

@@

We now move onto the case in which $N_0$ contains at least one inner vertex. If $N_0$ contains a face bounded by two parallel edge $e$ and $e'$ with $e\in E(F)$, then we can apply the inductive hypothesis on the cycle $F'$ obtained from $F$ by replacing $e$ with $e'$. (This is a valid application of induction since $F'$ has fewer faces of $G_0$ in its interior than $F$ does.)  In this way, we may assume that each edge of $F$ bounds a face of $N_0$ that contains at least three vertices.

Our $H$-partition $\mathcal{P}$ will include $P_1,\dots,P_k$ and a (possibly empty) part $S$ that will be determined by three vertices $v_1v_2v_3$ that belong to a common inner face $\tau$ of $N_0$.  We first explain how $v_1$, $v_2$ and $v_3$ are chosen and then explain how these are used to define $S$.  There is one case to consider for each possible value of $k$ (see \cref{boring_figure}):

@@

    \item If $k=3$ then assign each vertex $v$ of $N_0$ a colour $\alpha(v)\in\{1,\ldots,3\}$ as follows:  Let $w$ be the first vertex of $V(F)$ encountered on the path in $T$ from $v$ to the root of the tree that contains $v$.  (This implies that $w=v$ when $v$ is a vertex of $F$.)  Since $w\in V(F)$, $w\in P_i$ for some $i\in\{1,\ldots,k\}$ and we define $\alpha(v):=i$.  Now, for each $d'\in\{4,\ldots,d\}$ and each $d'$-sided inner face $F$ of $N_0$, add $d'-3$ edges inside of $F$ to split it into $3$-sided faces.  This yields a plane supergraph $N_0'$ of $N_0$ in which each inner face has exactly three edges on its boundary. By Sperner's Lemma (see \citep{Proofs4}) there exists an inner face $\tau':=v_1,v_2,v_3$ of $N_0'$ such that $\alpha(v_i)=i$ for each $i\in\{1,2,3\}$.\footnote{Although Sperner's Lemma is usually stated for simple graphs, the proof considers a certain graph defined over the faces of $N_0'$.  The proof relies only on the fact that the number of odd-degree vertices in this graph is even (known as the \defin{Handshake Lemma}), which is true even when this graph is non-simple.}  Let $\tau:=v_1,\ldots,v_p$ be  the facial cycle of $N_0$ that contains $\tau'$.
  \end{compactenum}

@@

  We now define the vertices in $S$.  For each $i\in\{1,2,3\}$, let $Q_i$ be the shortest path, in $T$, from $v_i$ to a vertex in $V(F)$.  Let $\overline{S}$ denote the subgraph of $N_0$ consisting of vertices and edges of $Q_1$, $Q_2$, $Q_3$, and $\tau$. Let $S:=V(\overline{S})\setminus V(F)$, let $Y:=(V(Q_1)\cup V(Q_2)\cup V(Q_3))\setminus V(F)$, and let $X:=S\setminus Y$. Observe that $|V(Q_i)\cap L_j|\le 1$   for each $i\in\{1,2,3\}$ and each $j\ge 0$.  Therefore $|Y\cap L_j|\le 3$ for each $j\ge 0$. Furthermore, $X\subseteq V(\tau)\setminus\{v_1,v_2,v_3\}$, so $|X|\le p-3\le d-3$. Therefore, using $S:=Y\cup X$ as a part in the partition $\PP$ satisfies condition~(ii).

@@

\begin{proof}[Proof of \cref{d_framed_product_stucture}]
Let $G$ be a $d$-framed multigraph with frame $G_0$ having outer face $F_0$
Let $T$ be a BFS forest of $G_0$ rooted at $V(F_0)$.  For each integer $j\ge 0$, let $L_j:=\{w\in V(G_0):\dist_G(w,V_0)=j\}$, so that $\langle L_0,L_1,\ldots\rangle$ is a layering of $G_0$.  Let $P_1:= V(F_0)$, let $Y_1$ be any subset of $\min\{|P_1|,3\}$ vertices in $P_1$ and let $X_1:=P_1\setminus Y_1$.  Then $G$, $G_0$, $T$, $F=F_0$, $k=1$, and $P_1=Y_1\cup X_1$ satisfy the conditions of \cref{induction}.  Apply \cref{induction} to obtain an $H$-partition $\mathcal{P}:=\{S_x:x\in V(H)\}$ of $G$ that satisfies Conditions~(i), (ii), and (iii).  Since any graph having a $3$-simple tree-decomposition is planar \cite{knauer.ueckerdt:simple,kratochvil.vaner:note,elmallah.colbourn:on}, Condition~(iii) implies that $H$ is planar.

Define $\mathcal{L}=\langle L_0',L_1'\ldots\rangle$ where $L_i'=L_{\lfloor d/2\rfloor i}\cup \cdots \cup L_{\lfloor d/2\rfloor(i+1)-1}$ for each integer $i\ge 0$. This is a layering of $G$ since $\dist_{G_0}(v,w)\le \floor{ \frac{d}{2} }$ for each edge $vw\in E(G)$.  By Condition~(ii), $|L_j\cap Y_x|\le 3$ for each integer $j\ge 0$, so $|L_i'\cap Y_x|\le 3\floor{\frac{d}{2}}$. Therefore, $|L_i'\cap S_x|= |L_i'\cap Y_x|+ |L_i'\cap X_x|\le \floor{ \frac{d}{2}} + d -3$ for each integer $i\ge 0$ and each $x\in V(H)$. The result now follows from \cref{PartitionProduct}.
\end{proof}

### 2022-11-07 response2.tex
Changed response. Remembered that some referee didn't like partition
Pfft
A bunch of changes

@@

\begin{response}
We do not understand this comment. Why can't there be a product structure
theorem for string graphs?
\end{response}

@@

\begin{response}
  We have replaced \\
  ``$P := (S_x : x \in V (T))$ is a sequence of pairwise disjoint sets that cover $V(G)$'' \\
  with \\
  ``$\mathcal{P}:=\{S_x : x\in V(T)\}$ is a partition of $V(G)$.''
\end{response}

@@

@@

@@

@@

@@

@@

@@

@@

15. page 19, second to last paragraph. Add a sentence about why such
a forest exists. The definition of a BFS layering is missing. A reader
may expect that you will just use the assumptions about H and P
that are stated, so it would be better to say something like “This gives
an H-partition... and conditions (i), (ii), and (iii) of Lemma 24 are
satisfied”.

>>>>

\begin{response}
  We have added a footnote here explaining that the proof of Sperner's Lemma is based on a certain graph defined on the faces of the embedding (triangles plus outer face).  It relies only on the fact that the number of odd-degree vertices in this graph is even, which is true even if this graph is not simple.
\end{response}

@@

\begin{response}
  We have added the qualifier "bounded-degree" here.
\end{response}

@@

\begin{response}
  We have replaced ``sequence'' with ``collection''.
\end{response}

@@

\begin{response}
  As part of addressing the referee's primary comment, we have introduced an extra claim that which shows that, for any $v\in V(G)$, there is a path in $H$ from $v$ to $a(v)$ of length at most $k-1$ in which $v$ participates in every vertex on this path, including $a(v)$.  We have rephrased this confusing sentence and replaced it with a reference to this claim.
\end{response}

@@

\begin{response}
  We have added a note at the beginning of Section 2.2 stating that, rather write ``$x$ is the unique vertex in $V(H)$ such that $v\in B_x$'' we write
  ``$P_x$ is the part in $\mathcal{H}$ that contains $v$.''  (Where the $H$-partition $\mathcal{H}:=\{B_x:\in V(H)\}$ has already been defined.)
\end{response}

@@

\begin{response}
  We have adjusted the definition of $C_x$ so that it only includes $x$ if $x\in V(J)$.  We agree that the overloading of the notation $T[\cdot]$ is slightly unfortunate, but can live with it since the two uses can be distinguished from the type of the argument.  For a single node $a$, $T[a]$ denote the subtree induced by all nodes whose bags contain $a$. For a set $S$, $T[S]$ is the subgraph of $T$ induced by $S$.
\end{response}

@@

\begin{response}
  We now specify that this forest contains one tree rooted at each vertex of $V_0$.
\end{response}

@@

\begin{response}
  We have added a couple of sentences here pointing to the exact lemma (Lemma~17) in the JACM paper and discussing the differences between the tripods here and there.
\end{response}

@@

\begin{response}
  We have added an explanation, prior to this, of how to eliminate the possibility that an edge of $F$ bounds a face with with only two vertices and two parallel edges.  Simply apply induction on the cycle $F'$ that uses the second of the parallel edges rather than the first.  This is valid since $F'$ has fewer faces in its interior.
\end{response}

@@

\begin{response}
  We have replaced "appropriate" with the "root of tree in $T$ that contains $v$".
\end{response}

@@

15. page 19, second to last paragraph. Add a sentence about why such
a forest exists. The definition of a BFS layering is missing.

\begin{response}
  The more precise definition of ``BFS forest rooted $V_0$'' takes care of this.
\end{response}

A reader may expect that you will just use the assumptions about H and P that are stated, so it would be better to say something like “This gives an H-partition... and conditions (i), (ii), and (iii) of Lemma 24 are satisfied”.
